A generalization of the Bradley-Terry model for draws in chess

### 2020-09-30 inst/jebo_chess_revision2.tex
write up rest monster

>>>>

\definecolor{DarkGreen}{RGB}{14,159,4}

\textcolor{DarkGreen}{It is reasonable to suppose that collusion of
  the type considered here is ``effective'' in the sense that
  noncollusive games are detrimental to future play in a way that
  collusively drawn games are not.  At the end of the tournament,
  Petrosian, Keres and Geller had each played a total of 18
  noncollusive matches while Fischer, Benko and Filip had played a
  total of 26 or 27 (Tal, retiring mid tournament due to illness,
  played only 21).  For each game in the tournament, we count the
  total noncollusive games played up to that point for both the white
  and black player.  If, in that game, the number of previous
  noncollusive games for the white and black player differs by more
  than one, we posit a ``rest
  monster'' whose strength helps the player who has played fewer
  noncollusive games.}

\textcolor{DarkGreen}{
  Observing that the rest monster is
  active in 49 of the 105 games of the tournament, we may consider
  these 49 games and, discarding draws, test whether rested players
  actually win more often than expected; observations are given in
  Table~\ref{restmonsterfisher} .  We reject the null of rested
  players having the same strength as their more tired opponents
  (one-sided Fisher's exact test, $p\simeq 10^{-4}$).}

\begin{table}
\textcolor{DarkGreen}{
  \caption{Results for won games in Cura\c{c}ao 1962  between
      rested and unrested players    \label{restmonsterfisher}}
  \centering
  \fbox{%
    \begin{tabular}{l|cc}\\
      & White rested & Black rested \\ \hline
      White wins  & $9$   & $ 3$   \\
      Black wins  & $0$   & $13$   \\
  \end{tabular}}
  }
\end{table}

\textcolor{DarkGreen}{The Bradley-Terry model together with a white
  player advantage, draw monster, and rest monster, estimates the
  strength of the rest monster at about 26\% (compare the draw monster
  at 24\% and the White advantage at about 3\%).  The null of the rest
  monster having zero strength is rejected with a likelihood ratio of
  about $e^{11.9}$ or a $p$-value of about $10^{-6}$.  It is an
  interesting observation that neither Table~\ref{restmonsterfisher},
  nor reified Bradley-Terry, admit a White advantage (one-sided
  McNemar test, $p=\frac{1}{8}$, likelihood ratio $1.609$
  respectively).  }
